
---
title: "Reconsidering how land titling could affect conflict between indigenous and non-indigenous peasants in Bolivia"
author:
  - Lorenza Fontana
  - Matias López
  - Huáscar Pacheco
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
  word_document: default
bibliography: /Users/lopezmar/Documents/Bolivia Land Reform/ref_land.bib
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  cache   = FALSE
)

```

```{r libraries}
library(tidyverse)
library(glmmTMB)
library(sjPlot)
library(broom.mixed)
library(ggplot2)
library(patchwork)
library(sf)
library(geodata)
library(stringdist)
library(brms)
library(posterior)
library(bayesplot)
```



```{r load-data}
bolivia <- read_csv(
  "https://raw.githubusercontent.com/Matias-Lopez-13/Bolivia-Land-Reform-Project/main/bolivia_updated_TIOC.csv"
)
```

```{r clean-data}



library(dplyr)
library(stringi)

bolivia <- bolivia %>%
  mutate(
    Municipality_clean = case_when(
      Municipality == "San Antonio De Esmoruco"  ~ "San Antonio de Esmoruco",
      Municipality == "San Pedro De Curahuara"   ~ "San Pedro de Curahuara",
      Municipality == "Curahuara De Carangas"    ~ "Curahuara de Carangas",
      Municipality == "Cruz De Machacamarca"     ~ "Cruz de Machacamarca",
      Municipality == "San Pedro De Tiquina"     ~ "San Pedro de Tiquina",
      Municipality == "Nazacara De Pacajes"      ~ "Nazacara de Pacajes",
      Municipality == "San Pedro De Quemes"      ~ "San Pedro de Quemes",
      Municipality == "Santiago De Machaca"      ~ "Santiago de Machaca",
      Municipality == "Santiago De Huata"        ~ "Santiago de Huata",
      Municipality == "Waldo Ballivián"          ~ "Waldo Ballivian",
      Municipality == "Colpa Belgica"            ~ "Colpa Bélgica",
      Municipality == "Puerto Suárez"            ~ "Puerto Suarez",
      Municipality == "Salinas De Garci Mendoza" ~ "Salinas de Garcí Mendoza",
      Municipality == "Santa Cruz De La Sierra"  ~ "Santa Cruz de la Sierra",
      Municipality == "San Antonio De Lomerío"   ~ "San Antonio de Lomerio",
      Municipality == "Ascensión De Guarayos"    ~ "Ascención de Guarayos",
      Municipality == "Yamparaez"                ~ "Yamparáez",
      Municipality == "Postrer Valle"            ~ "Postrervalle",
      Municipality == "Jesús De Machaca"         ~ "Jesús de Machaka",
      Municipality == "Sopachuy"                 ~ "Sopachui",
      Municipality == "Urubichá"                 ~ "Urubicha",
      Municipality == "Villamontes"              ~ "Villa Montes",
      Municipality == "Belén De Andamarca"       ~ "Belén de Andamarca",
      Municipality == "Choque Cota"              ~ "Choquecota",
      Municipality == "Bolívar"                  ~ "Bolivar",
      Municipality == "Vitichi"                  ~ "Vitiche",
      Municipality == "Zudáñez"                  ~ "Zudañez",
      Municipality == "Moro Moro"                ~ "Moromoro",
      Municipality == "Sica Sica"                ~ "Sicasica",
      Municipality == "Yunguyo De Litoral"       ~ "Yunguyo del Litoral",
      Municipality == "Carmen Rivero Tórrez"     ~ "El Carmen Rivero Tórrez",
      Municipality == "Gral Saavedra"            ~ "General Saavedra",
      Municipality == "Pari - Paria - Soracachi" ~ "Soracachi",
      Municipality == "Pto Carabuco"             ~ "Puerto Carabuco",
      Municipality == "San Andrés De Machaca"    ~ "San Andrés de Machaca",
      Municipality == "San Ignacio De Velasco"   ~ "San Ignacio de Velasco",
      Municipality == "San José De Chiquitos"    ~ "San José de Chiquitos",
      Municipality == "San Juan De Yapacaní"     ~ "San Juan del Yapacaní",
      Municipality == "San Miguel De Velasco"    ~ "San Miguel de Velasco",
      Municipality == "San Pablo De Lipez"       ~ "San Pablo de Lípez",
      Municipality == "San Pedro De Totora"      ~ "San Pedro de Totora",
      Municipality == "Santa Ana De Yacuma"      ~ "Santa Ana de Yacuma",
      Municipality == "Santa Rosa Del Sara"      ~ "Santa Rosa del Sara",
      Municipality == "Santiago De Andamarca"    ~ "Santiago de Andamarca",
      Municipality == "Sp De Buena Vista"        ~ "San Pedro de Buena Vista",
      Municipality == "Villa Nueva (Loma Alta)"  ~ "Villa Nueva",
      TRUE ~ Municipality
    ),

    Municipality_clean = stri_trans_general(Municipality_clean, "Latin-ASCII"),
    Municipality_clean = toupper(trimws(Municipality_clean))
  )

```


## Introduction

The policy framework for agrarian reforms in Latin America seems to have shifted from a focus on class conflict between lords and peasants [@albertus2015; @Lenin1899CapitalismRussia; @Ondetti2010; @wolf1969peasantwars] towards a focus on conflict between historically discriminated ethnic groups and the state [@Dyzenhaus2026; @Fontana_book; @Ngcukaitobi2018]. While indigenous peoples and descendants of enslaved peoples have benefited from this new policy framework, scholars have asked if the shift from class to race could have unintended negative consequences.

Based on the qualitative case study of the municipality of Apolo in Bolivia, Fontana [-@Fontana2014;-@Fontana2015] argued that the identity criterion can foster conflict within the peasantry. The study shows that unionized non-indigenous peasants in Apolo resented what they perceived as unfair advantages of indigenous groups. The latter received land titles from Bolivia's more recent and highly ethnicized agrarian reform program. Thus, the vessel for conflict resolution between indigenous peoples and the state inadvertently generated new grounds for conflict.

The mechanism linking land titling and conflict among peasants seems clear within the case of Apolo, but is such mechanism generalizable to the Bolivian agrarian reform more broadly? While it is hard to answer this question definitively, we provide a framework for testing a descriptive hypothesis that associates the recognition of ancestral land with conflict.

Until now it was hard to estimate the association between land titling and conflict because standardized nationwide data on conflict in Bolivia were not available. We partnered with Fundación UNIR Bolivia (www.unirbolivia.org), which uses national and regional media sources to take note of protests in most Bolivian municipalities. Using protest as a proxy for conflict, we were able to map these events and match them with the creation of new ethic-based territories. These are known as Native Communal Territories (*Tierras Comunitarias de Origen*, TCOs), renamed Indigenous Native Peasant Territories (*Territorios Indígena Originario Campesinos*, TIOCs) in 2010. Hereafter, we refer to them as TCO/TIOCs to reflect both designations across the period covered by our data (2006–2018). 

Our general goal was to test if countrywide data would be consistent with the causal processes that seems to have taken place in the municipality of Apolo, formalized as our main hypothesis $H_1$: *Conflict follows titling*. Testing $H_1$ implies working with two rival hypotheses. The most straightforward is the sharp null, $H_0$: *Land titling has no effect on posterior conflict*. The second rival states the opposite, $H_2$ *Land titling reduces posterior conflict*. Naturally, the prior expectation supports this second rival hypothesis because it seems safe to assume that governments implement recognition policies under the idea that these would solve, not cause, conflict.

## What we know about conflict among peasants

Research on other regions (mainly the Middle East and Africa) suggest that ethnic peasant conflicts can challenge social stability when they mobilize ethnic divides [@gurr1993], which is not how Latin American cases are usually interpreted [@gurr1995].

However, scholars have argued that race and ethnicity gained political salience as authoritarian regimes that had previously suppressed ethnic mobilization gave way to democracy, opening the door to competitive politics and a more prominent role for local organizations [@yashar2005; @andolina2009; @rousseau_dargent_2019].

Many authors --mostly in economics-- associate violent conflict and ethnic diversity [@collier2004; @miguel_satyanath_sergenti_2004; @montalvo_reynal_2005], and Latin America is certainly diverse. This type of conflict can be understood as an intense emotional reaction to grievances driven by ethnic cleavages [@alesina1999; @alesina2005].

Others have emphasized that ethnic identities and cleavages are more likely shaped by politics rather than the other way around [@chandra2012; @singh_vomhau_2016; @wimmer_2013]. For instance, researchers have examined how the institutionalization of ethnic divisions—such as through censuses or official ethnic certifications—can create competitive dynamics that increase the risk of escalating conflict [@lieberman_singh_2012;@lieberman_singh_2017]. Therefore, the institutional context within which ethnic practices occur may be the most critical factor [@wimmer_2013].

Overall, the literature offers limited information about whether we should expect conflict to follow the TCO program in Bolivia or not. 

## Our data

We have data about land titling and conflict covering the 2004-2018 window, the distribution can be seen in Figure 1. 




```{r Figure 1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Distribution of land titles and conflict", fig.align='center', out.width='80%', fig.width=9, fig.height=6}

library(dplyr)
library(ggplot2)
library(patchwork)

# ----------------------------------------------------------
# Annual totals
# ----------------------------------------------------------

annual_totals <- bolivia %>%
  filter(Year >= 2004, Year <= 2018) %>%
  group_by(Year) %>%
  summarise(
    total_titles = sum(Title_check, na.rm = TRUE),
    total_protests = sum(N_protests, na.rm = TRUE),
    .groups = "drop"
  )

# ----------------------------------------------------------
# Land titles
# ----------------------------------------------------------

p_titles_bar <- ggplot(
  annual_totals,
  aes(x = Year, y = total_titles)
) +
  geom_col(
    fill = "#6FA9C4",
    width = 0.7
  ) +
  geom_text(
    aes(label = total_titles),
    vjust = -0.5,
    size = 2.5,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = 2004:2018
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Total land titles issued per year",
    x = NULL,
    y = "Number of titles"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "plain"
    ),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

# ----------------------------------------------------------
# Protests
# ----------------------------------------------------------

p_protests <- ggplot(
  annual_totals,
  aes(x = Year, y = total_protests)
) +
  geom_col(
  fill = "#A95A45",
  width = 0.7
) +
  geom_text(
    aes(label = total_protests),
    vjust = -0.5,
    size = 2.5,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = 2004:2018
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Total protests per year",
    x = NULL,
    y = "Number of protests"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "plain"
    ),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

# ----------------------------------------------------------
# Combine
# ----------------------------------------------------------

p_titles_bar / p_protests +
  plot_layout(heights = c(1, 1))

```


```{r country-years}
library(dplyr)

# total country-years in full sample
total_country_years <- bolivia %>%
  summarise(n = n_distinct(paste(Municipality, Year))) %>%
  pull(n)

# restricted “good data” period
good_country_years <- bolivia %>%
  filter(Year >= 2006, Year <= 2018) %>%
  summarise(n = n_distinct(paste(Municipality, Year))) %>%
  pull(n)

# extension period
extension_country_years <- bolivia %>%
  filter(Year >= 2011, Year <= 2018) %>%
  summarise(n = n_distinct(paste(Municipality, Year))) %>%
  pull(n)
```

The dataset contains `r total_country_years` municipality-years. Figure 2 shows the spatial distribution of TCO/TIOCs and protests.





```{r Figure 2, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Spacial distribution of TCO/TIOCs and protests in Bolivia", fig.align='center', out.width='80%'}

# Download Bolivia municipality shapefile (level 3 = municipalities)
bol_shp <- gadm("BOL", level = 3, path = tempdir()) %>%
  st_as_sf()




# Common fixes for Bolivia — accents, abbreviations
bolivia <- bolivia %>%
  mutate(Municipality = Municipality %>%
           stringr::str_to_title() %>%
           stringr::str_replace("San ", "San ") %>%   # adjust as needed
           stringr::str_trim())


# Summarize — total titles and total protests per municipality
bol_summary <- bolivia %>%
  group_by(Municipality) %>%
  summarise(
    total_titles   = sum(Title_check, na.rm = TRUE),
    total_protests = sum(N_protests, na.rm = TRUE),
    any_titled     = as.integer(any(Title_check > 0)),
    mean_indigenous = mean(indigenous_percent2012, na.rm = TRUE)
  )

# Join to shapefile
bol_map <- bol_shp %>%
  left_join(bol_summary, by = c("NAME_3" = "Municipality"))

my_pal <- c("#87CEEB", "#C0392B")

apolo <- bol_map %>%
  filter(NAME_3 == "Apolo") %>%
  st_centroid() %>%
  st_coordinates() %>%
  as.data.frame()

p_titles_map <- ggplot(bol_map) +
  geom_sf(aes(fill = total_titles), color = "white", linewidth = 0.1) +
  scale_fill_gradient(low = my_pal[1], high = my_pal[2],
                      na.value = "grey85", name = "Titles") +
  annotate("segment",
           x = apolo$X + 2, y = apolo$Y + 2,
           xend = apolo$X, yend = apolo$Y,
           arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
           color = "black", linewidth = 0.6) +
  annotate("text", x = apolo$X + 2.1, y = apolo$Y + 2.2,
           label = "Apolo", size = 3, hjust = 0, fontface = "bold") +
  labs(title = "Land titles issued") +
  theme_void(base_size = 11) +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"))

p_protests_map <- ggplot(bol_map) +
  geom_sf(aes(fill = total_protests), color = "white", linewidth = 0.1) +
  scale_fill_gradient(low = my_pal[1], high = my_pal[2],
                      na.value = "grey85", name = "Protests") +
  annotate("segment",
           x = apolo$X + 2, y = apolo$Y + 2,
           xend = apolo$X, yend = apolo$Y,
           arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
           color = "black", linewidth = 0.6) +
  annotate("text", x = apolo$X + 2.1, y = apolo$Y + 2.2,
           label = "Apolo", size = 3, hjust = 0, fontface = "bold") +
  labs(title = "Protests") +
  theme_void(base_size = 11) +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"))

p_titles_map + p_protests_map
```

Our intuition was to treat these as count data, i.e. count the number of titles and the number of protests based on a model to test if both counts are associated. Because our test implies the contrast between two different theories about the effect of land titles in Bolivia, we decided for two distinct approaches.

First, we use zero-inflated negative binomial models which allows us to account for violations of the assumptions of a Poisson distribution. The conditional count component models the expected number of protests as a function of land titles issued, indigenous population percentage, poverty rate, population growth, internal migration, illiteracy rate, forested area, cultivated area, travel time to the departmental capital, municipal area, working-age population share, and the share of population living outside urban centers. All time-invariant controls are drawn from the 2012 national census, which is the closest in time to any given year in the dataset.

To account for the nested structure of the data — municipalities observed repeatedly over time — the model includes municipality-level random intercepts, allowing baseline protest propensity to vary across units independently of the covariates. A second specification adds department-level random intercepts with municipalities nested within departments, accounting for the possibility that unobserved regional factors (INRA office capacity, departmental politics, geographic clustering of land conflict) jointly predict titling and protest. Coefficients are reported as incidence rate ratios (IRR), where values below 1 indicate a negative association between the predictor and protest frequency. Standard errors are clustered at the municipality level.

Second, we use a Bayesian version of these same models to more clearly contrast the three competing theories: TCO/TIOCs positively affect protest vs TCO/TIOCs do not affect protest vs TCO/TIOCs negatively affect protest. The Bayesian model also accounts for overdispersed count data and includes municipality-level random intercepts to capture unobserved heterogeneity across units, while controlling for key socioeconomic covariates such as poverty, demographics, land use, and accessibility. Weakly informative priors are used to regularize parameter estimates, and posterior distributions are summarized to assess both the magnitude and directional probability of the effect.

We inform the prior belief in $H_2$ using a sensitivity analysis approach, i.e. a range of different prior likelihoods which reflect mild to strong negative effects of land titling on conflict.

## Results



```{r models, results='hide'}
library(dplyr)
library(glmmTMB)
library(sjPlot)
library(broom.mixed)
library(ggplot2)

# ----------------------------------------------------------
# 1. Recreate lags using the full municipal panel
# ----------------------------------------------------------

bolivia <- bolivia %>%
  arrange(Municipality, Year) %>%
  group_by(Municipality) %>%
  mutate(
    Title_lag1 = if_else(
      Year - lag(Year, 1) == 1,
      as.numeric(lag(Title_check, 1)),
      NA_real_
    ),
    Title_lag2 = if_else(
      Year - lag(Year, 2) == 2,
      as.numeric(lag(Title_check, 2)),
      NA_real_
    ),
    Title_lag3 = if_else(
      Year - lag(Year, 3) == 3,
      as.numeric(lag(Title_check, 3)),
      NA_real_
    )
  ) %>%
  ungroup()

# ----------------------------------------------------------
# 2. Main analytical sample: 2006-2018
# ----------------------------------------------------------

dat <- bolivia %>%
  filter(Year >= 2006, Year <= 2018)

# ----------------------------------------------------------
# 3. Baseline model
# ----------------------------------------------------------

m_baseline <- glmmTMB(
  N_protests ~ Title_check +
    indigenous_percent2012 +
    poor_percent2012 + popgrowth_per_year +
    migration_internal + literacy_percent_no +
    forested_area_percent + area_total_cultivated +
    time_to_capital + area_municip +
    working_age_percent_yes + pop_percent_outside +
    (1 | Municipality),
  ziformula = ~ poor_percent2012,
  family = nbinom2,
  data = dat
)

# Department models omitted because the original
# specifications were identical to the baseline.

# ----------------------------------------------------------
# 4. Common sample for lag comparisons
# ----------------------------------------------------------

model_vars <- c(
  "N_protests", "Title_check",
  "Title_lag1", "Title_lag2", "Title_lag3",
  "indigenous_percent2012", "poor_percent2012",
  "popgrowth_per_year", "migration_internal",
  "literacy_percent_no", "forested_area_percent",
  "area_total_cultivated", "time_to_capital",
  "area_municip", "working_age_percent_yes",
  "pop_percent_outside", "Municipality"
)

dat_lags <- dat %>%
  filter(
    if_all(all_of(model_vars), ~ !is.na(.x))
  )

# ----------------------------------------------------------
# 5. Models with different lag structures
# ----------------------------------------------------------

m_lag0 <- glmmTMB(
  N_protests ~ Title_check +
    indigenous_percent2012 +
    poor_percent2012 + popgrowth_per_year +
    migration_internal + literacy_percent_no +
    forested_area_percent + area_total_cultivated +
    time_to_capital + area_municip +
    working_age_percent_yes + pop_percent_outside +
    (1 | Municipality),
  ziformula = ~ poor_percent2012,
  family = nbinom2,
  data = dat_lags
)

m_lag1 <- glmmTMB(
  N_protests ~ Title_check + Title_lag1 +
    indigenous_percent2012 +
    poor_percent2012 + popgrowth_per_year +
    migration_internal + literacy_percent_no +
    forested_area_percent + area_total_cultivated +
    time_to_capital + area_municip +
    working_age_percent_yes + pop_percent_outside +
    (1 | Municipality),
  ziformula = ~ poor_percent2012,
  family = nbinom2,
  data = dat_lags
)

m_lag2 <- glmmTMB(
  N_protests ~ Title_check + Title_lag2 +
    indigenous_percent2012 +
    poor_percent2012 + popgrowth_per_year +
    migration_internal + literacy_percent_no +
    forested_area_percent + area_total_cultivated +
    time_to_capital + area_municip +
    working_age_percent_yes + pop_percent_outside +
    (1 | Municipality),
  ziformula = ~ poor_percent2012,
  family = nbinom2,
  data = dat_lags
)

m_lag3 <- glmmTMB(
  N_protests ~ Title_check + Title_lag3 +
    indigenous_percent2012 +
    poor_percent2012 + popgrowth_per_year +
    migration_internal + literacy_percent_no +
    forested_area_percent + area_total_cultivated +
    time_to_capital + area_municip +
    working_age_percent_yes + pop_percent_outside +
    (1 | Municipality),
  ziformula = ~ poor_percent2012,
  family = nbinom2,
  data = dat_lags
)

# ----------------------------------------------------------
# 6. Regression table
# ----------------------------------------------------------

tab_model(
  m_lag0, m_lag1, m_lag2, m_lag3,
  transform   = "exp",
  show.icc    = FALSE,
  show.re.var = FALSE,
  dv.labels   = c("Lag 0", "Lag 1", "Lag 2", "Lag 3"),
  pred.labels = c(
    Title_check             = "Land titles (t)",
    Title_lag1              = "Land titles (t-1)",
    Title_lag2              = "Land titles (t-2)",
    Title_lag3              = "Land titles (t-3)",
    indigenous_percent2012  = "Indigenous %",
    poor_percent2012        = "Poverty %",
    popgrowth_per_year      = "Population growth",
    migration_internal      = "Internal migration",
    literacy_percent_no     = "Illiteracy %",
    forested_area_percent   = "Forested area %",
    area_total_cultivated   = "Cultivated area",
    time_to_capital         = "Time to capital",
    area_municip            = "Municipality area",
    working_age_percent_yes = "Working age %",
    pop_percent_outside     = "Pop. outside urban"
  ),
  title = "Effect of land titling on protests by lag structure (IRR)"
)

# ----------------------------------------------------------
# 7. Prepare data for Figure 4
# ----------------------------------------------------------

lag_plot_data <- bind_rows(
  tidy(m_lag0, conf.int = TRUE, exponentiate = TRUE) %>%
    filter(
      effect == "fixed",
      component == "cond",
      term == "Title_check"
    ) %>%
    mutate(lag = "Same year"),

  tidy(m_lag1, conf.int = TRUE, exponentiate = TRUE) %>%
    filter(
      effect == "fixed",
      component == "cond",
      term == "Title_lag1"
    ) %>%
    mutate(lag = "One year after"),

  tidy(m_lag2, conf.int = TRUE, exponentiate = TRUE) %>%
    filter(
      effect == "fixed",
      component == "cond",
      term == "Title_lag2"
    ) %>%
    mutate(lag = "Two years after"),

  tidy(m_lag3, conf.int = TRUE, exponentiate = TRUE) %>%
    filter(
      effect == "fixed",
      component == "cond",
      term == "Title_lag3"
    ) %>%
    mutate(lag = "Three years after")
) %>%
  mutate(
    lag = factor(
      lag,
      levels = c(
        "Same year",
        "One year after",
        "Two years after",
        "Three years after"
      )
    ),
    significant = p.value < 0.1
  )

# ----------------------------------------------------------
# 8. Objects for automatic reporting in the text
# ----------------------------------------------------------

# Contemporaneous association
title_coef <- fixef(m_lag0)$cond["Title_check"]
title_irr <- exp(title_coef)

title_p <- tidy(m_lag0) %>%
  filter(
    effect == "fixed",
    component == "cond",
    term == "Title_check"
  ) %>%
  pull(p.value)

# Percentage change in expected protests
title_pct_change <- (title_irr - 1) * 100

title_pct <- abs(title_pct_change)

title_direction <- ifelse(
  title_pct_change < 0,
  "lower",
  "higher"
)

# One-year lag association
lag1_coef <- fixef(m_lag1)$cond["Title_lag1"]
lag1_irr <- exp(lag1_coef)

lag1_p <- tidy(m_lag1) %>%
  filter(
    effect == "fixed",
    component == "cond",
    term == "Title_lag1"
  ) %>%
  pull(p.value)

lag1_pct_change <- (lag1_irr - 1) * 100

lag1_pct <- abs(lag1_pct_change)

lag1_direction <- ifelse(
  lag1_pct_change < 0,
  "lower",
  "higher"
)

# ----------------------------------------------------------
# 9. Verify sample sizes
# ----------------------------------------------------------

sample_check <- data.frame(
  Model = c(
    "Baseline",
    "Lag 0", "Lag 1", "Lag 2", "Lag 3"
  ),
  Observations = c(
    nobs(m_baseline),
    nobs(m_lag0),
    nobs(m_lag1),
    nobs(m_lag2),
    nobs(m_lag3)
  )
)

print(sample_check)

```







```{r Figure 4, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Estimated association between TCO/TIOCs and protests over time (2006–2018)", fig.align='center', out.width='80%', fig.width=9, fig.height=5}

library(ggplot2)

ggplot(
  lag_plot_data,
  aes(
    x = lag,
    y = estimate,
    ymin = conf.low,
    ymax = conf.high
  )
) +
  geom_hline(
    yintercept = 1,
    linetype = "dashed",
    color = "grey50"
  ) +
  geom_pointrange(
    color = "#6FA9C4",
    size = 0.8,
    linewidth = 0.8
  ) +
  scale_y_continuous(
    breaks = seq(0, 2, by = 0.25)
  ) +
  coord_cartesian(ylim = c(0.5, 1.5)) +
  labs(
    x = NULL,
    y = "Incidence Rate Ratio (IRR)",
    caption = "IRR < 1 indicates fewer expected protests. Error bars represent 95% confidence intervals."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

```




Figure 4 shows the estimated effect of land titles. Models estimate the association between land titling and protests in the same year and up to three years afterward, using municipality-year observations from 2006 to 2018. Estimates suggest that land titling did not reduce the number of protests in time. 

### Bayesian measures

We informed the same model with priors that reflect the most intuitive hypothesis of conflict reduction, $H_2$. This means that we estimate how more likely $H_1$ becomes if we assume that titling should reduce conflict. Bayesian models allow us to estimate an updated degree of belief in each theory, including the null hypothesis. Table 1 describes the statistics.


```{r bayesian-models, results='hide', cache=TRUE}

library(brms)
library(posterior)
library(dplyr)
library(tibble)

dat <- bolivia %>%
  filter(Year >= 2006, Year <= 2018)

# ----------------------------------------------------------
# Common priors
# ----------------------------------------------------------

base_priors <- c(
  prior(normal(0, 1),    class = b),
  prior(normal(0, 10),   class = Intercept),
  prior(cauchy(0, 2.5),  class = sd),
  prior(gamma(0.1, 0.1), class = shape)
)

# ----------------------------------------------------------
# Model formula
# ----------------------------------------------------------

f <- bf(
  N_protests ~ Title_check +
    indigenous_percent2012 + poor_percent2012 + popgrowth_per_year +
    migration_internal + literacy_percent_no + forested_area_percent +
    area_total_cultivated + time_to_capital + area_municip +
    working_age_percent_yes + pop_percent_outside +
    (1 | Municipality)
)

# ----------------------------------------------------------
# H0: No effect
# Prior centered at zero
# ----------------------------------------------------------

m_H0 <- brm(
  f,
  family = negbinomial(),
  data = dat,
  prior = c(
    base_priors,
    prior(normal(0, 1), class = b, coef = "Title_check")
  ),
  chains = 4, iter = 2000, warmup = 1000,
  cores = 4, seed = 42,
  control = list(adapt_delta = 0.95)
)

# ----------------------------------------------------------
# H1: Land titling increases conflict
# Prior centered on a positive effect
# ----------------------------------------------------------

m_H1 <- brm(
  f,
  family = negbinomial(),
  data = dat,
  prior = c(
    base_priors,
    prior(normal(0.3, 0.5), class = b, coef = "Title_check")
  ),
  chains = 4, iter = 2000, warmup = 1000,
  cores = 4, seed = 42,
  control = list(adapt_delta = 0.95)
)

# ----------------------------------------------------------
# H2: Land titling reduces conflict
# Prior centered on a negative effect
# ----------------------------------------------------------

m_H2 <- brm(
  f,
  family = negbinomial(),
  data = dat,
  prior = c(
    base_priors,
    prior(normal(-0.3, 0.5), class = b, coef = "Title_check")
  ),
  chains = 4, iter = 2000, warmup = 1000,
  cores = 4, seed = 42,
  control = list(adapt_delta = 0.95)
)

```


```{r table1}
library(dplyr)
library(tibble)
library(posterior)
library(knitr)

theory_probs <- function(model, rope = 0.05) {

  post <- as_draws_df(model)

  tibble(
    `H0: No Effect` = paste0(
      round(mean(abs(post$b_Title_check) <= rope) * 100, 1), "%"
    ),
    `H1: Increase Protests` = paste0(
      round(mean(post$b_Title_check > rope) * 100, 1), "%"
    ),
    `H2: Reduce Protests` = paste0(
      round(mean(post$b_Title_check < -rope) * 100, 1), "%"
    )
  )
}

table1_data <- bind_rows(
  "Neutral prior"  = theory_probs(m_H0),
  "Positive prior" = theory_probs(m_H1),
  "Negative prior" = theory_probs(m_H2),
  .id = "Prior specification"
) %>%
  select(
    `Prior specification`,
    `H0: No Effect`,
    `H1: Increase Protests`,
    `H2: Reduce Protests`
  )

knitr::kable(
  table1_data,
  caption = "Posterior probabilities of hypotheses given data",
  align = c("l", "c", "c", "c")
)
```



None of the hypotheses are highly likely, however the data are more expected given $H_1$. But the probability of both rivals is not sufficiently low for rejecting either theory. In the next section we discuss what these new information may say about the case of Apolo and about the possible external validity of the mechanisms estimated for explaining conflict there.

### The Apolo municipality in face of new evidence


The conflict in Apolo emerged during the late 1990s when a faction of the local peasant federation broke away to form the indigenous organization CIPLA (Central Indígena del Pueblo Leco). The dispute centered on the recognition of a Leco indigenous identity and the creation of the TCO-Leco indigenous territory. While indigenous activists viewed the process as the recovery of ancestral rights and territory, many unionized peasants saw it as an illegitimate attempt to gain privileged access to land and political influence. Tensions escalated as communities shifted from the peasant federation to CIPLA and as the TCO titling process advanced. 

The conflict reached its peak in 2007, after the first TCO-Leco title was issued, leading to marches, blockades, hunger strikes, occupations of public buildings, and the invasion of part of Madidi National Park. Although negotiations reduced tensions, the dispute remained unresolved and continued to shape local politics and identities. Figure 5 describes this process.

```{r Figure 5, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Causal process in Apolo", fig.align='center', out.width='80%'}

library(tidyverse)

timeline <- tribble(
  ~year, ~event,
  1999, "TCO claim submitted",
  2003, "Cadastral study begins",
  2006, "TCO title issued",
  2007, "Mass mobilizations of non-indigenous peasants",
  2010, "Tribunal upholds TCO"
)


ggplot(timeline, aes(x = year, y = 0)) +
  geom_hline(yintercept = 0, linewidth = 0.8, color = "#87CEEB") +
  geom_point(size = 3, color = "#C0392B") +
  geom_text(aes(label = year),
            vjust = -1.2, fontface = "bold", size = 3.5) +
  geom_text(aes(label = event),
            angle = 45, hjust = 0,
            nudge_y = 0.15, size = 3.2) +
  scale_x_continuous(breaks = timeline$year) +
  coord_cartesian(ylim = c(-0.2, 0.6)) +
  labs(
    title = "Timeline of the Apolo Conflict",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank()
  )

```

Does the case of Apolo fit the pattern found in the quantitative dataset? It is certainly not incompatible with average estimates because the data, although inconclusive, make $H_1$ more expected. So models tell us that what was observed in Apolo is more likely if compared to contrasting scenarios. In other words, we should not be surprised by conflict emerging there. 



```{r bayesian-models-no-apolo, results='hide', message=FALSE, warning=FALSE}

library(dplyr)
library(brms)

# Analytical sample excluding Apolo
dat_no_apolo <- bolivia %>%
  filter(
    Year >= 2006,
    Year <= 2018,
    !is.na(Municipality),
    trimws(as.character(Municipality)) != "Apolo"
  )

# Model formula
f_apolo <- bf(
  N_protests ~ Title_check +
    indigenous_percent2012 + poor_percent2012 +
    popgrowth_per_year + migration_internal +
    literacy_percent_no + forested_area_percent +
    area_total_cultivated + time_to_capital +
    area_municip + working_age_percent_yes +
    pop_percent_outside +
    (1 | Municipality)
)

# Priors
priors_apolo <- c(
  prior(normal(0, 1), class = b),
  prior(normal(0, 1), class = b, coef = "Title_check"),
  prior(normal(0, 10), class = Intercept),
  prior(cauchy(0, 2.5), class = sd),
  prior(gamma(0.1, 0.1), class = shape)
)

# Estimate model
m_no_apolo <- brm(
  f_apolo,
  data = dat_no_apolo,
  family = negbinomial(),
  prior = priors_apolo,
  chains = 4,
  iter = 2000,
  warmup = 1000,
  cores = 4,
  seed = 42,
  control = list(adapt_delta = 0.95),
  file = "m_no_apolo",
  file_refit = "on_change"
)

```





```{r table-no-apolo}

library(posterior)
library(knitr)

# Extract posterior draws
beta <- as_draws_df(m_no_apolo)$b_Title_check

# Calculate posterior probabilities
table_no_apolo <- data.frame(
  `H0: No Effect` = mean(abs(beta) <= 0.05) * 100,
  `H1: Increase Protests` = mean(beta > 0.05) * 100,
  `H2: Reduce Protests` = mean(beta < -0.05) * 100,
  check.names = FALSE
)

# Format as percentages
table_no_apolo[] <- lapply(
  table_no_apolo,
  function(x) sprintf("%.1f%%", x)
)

knitr::kable(
  table_no_apolo,
  caption = "Posterior probabilities of coefficient regions excluding Apolo (2006–2018)",
  align = "ccc"
)

```


On the other hand, the data on other municipalities suggest no clear pattern. Which makes us think that other mechanisms may connect TCO/TIOCs with different outcomes.
 

# Conclusion

The literature on ethnic cleavages and agrarian reform is imprecise about what we should expect in terms of conflict after the implementation of recognition policies in Latin America. Bolivia is an important case because it represents a clear instance where land policies follow recognition politics quite strictly. 

Fontana [-@Fontana2014; -@Fontana2014] showed in great detail the machanism linking the implementation of the TCO program in Apolo triggered with conflict within the pesantry. Such mechanism tells us that the TCO mobilized non-indigenous unionized rural workers against beneficiaries. Here we use a panel of conflict in Bolivia and data on TCO/TIOCs to estimate if this pattern describes the Bolivian case more broadly.

Results indicate that this is more likely not a generalizable pattern. The data tell us that the agrarian reform program with a focus on ancestral land in not strongly associated with neither conflict resolution nor conflict increase. 

We are happy to provide the academic community with new data, which we could access thanks to NGO partners. However more data are needed to address this important question. These could include satellite images and social media data to better estimate how recognition policies shapes conflict.  


# References
